A Statistical Dynamical Model to Predict Extreme Events and Anomalous Features in Shallow Water Waves with Abrupt Depth Change

### 2018-11-25 paper1_tKdV/tKdV_theory.tex
NM

Recent laboratory experiments in weakly turbulent shallow water reveal a remarkable transition from Gaussian to anomalous behavior as surface waves cross an abrupt depth change (ADC). 
A normally-distributed incoming wave train, upstream of the ADC, transitions to a highly non-Gaussian outgoing wave train, downstream of the ADC, that exhibits large positive skewness of the surface height and more frequent extreme events \cite{bolles2018anomalous}.
Here we develop a statistical dynamical model to explain and quantitatively
predict the above anomalous statistical behavior as experimental control
parameters are varied. The first step is to use incoming and outgoing
truncated Korteweg-de Vries (TKdV) equations matched in time at the
ADC. The TKdV equation is a Hamiltonian system which induces incoming
and outgoing statistical Gibbs invariant measures which are statistically
matched at the ADC. The statistical matching of the known nearly Gaussian
incoming Gibbs state at the ADC completely determines the predicted
anomalous outgoing Gibbs state, which can be calculated by a simple
MCMC algorithm, verified by direct numerical simulations, and successfully
captures key features of the experiment. There is even an analytic
formula for the anomalous outgoing skewness. The strategy here should
be useful for predicting extreme anomalous statistical behavior in
other dispersive media in different settings.

@@

\section{Experiments showing anomalous wave statistics by the abrupt depth change}

A series of experiments are carried out \cite{bolles2018anomalous}
studying the anomalous statistical behaviors in surface water waves
going through an abrupt depth transition. 
The unidirectional waves propagate along a water tank over a step in the bottom topography,
and the surface displacements of the wave levels are measured at several
upstream and downstream locations. The wave field is excited by a
paddle wheel forcing with angle
\[
\theta\left(t\right)=\theta_{0}+\Delta\theta\sum_{n=1}^{N}a_{n}\cos\left(\omega_{n}t+\delta_{n}\right),\;E\sim\left(\Delta\theta\right)^{2}\sum_{n}a_{n}^{2}\omega_{n}^{2}.
\]
The total energy $E$ in the system is determined by the angle amplitude
$\Delta\theta$. Several observations are found in the anomalous wave
statistics:
\begin{itemize}
\item Distinct statistics are found between the incoming and outgoing wave
disturbances: the incoming waves display near-Gaussian statistics,
while the outgoing waves show skewness toward the positive displacement.
\item The non-Gaussian statistics is related with the total energy contained
in the system: larger driving amplitude $\Delta\theta$ will generate
stronger skewness in the PDFs.
\item The waves also show different characteristic peak wavelengths in incoming
and outgoing flows.
\end{itemize}

>>>>

Recent laboratory experiments in weakly turbulent shallow water reveal a remarkable transition from Gaussian to anomalous behavior as surface waves cross an abrupt depth change (ADC). 
A normally-distributed incoming wave train, upstream of the ADC, transitions to a highly non-Gaussian outgoing wave train, downstream of the ADC, that exhibits large positive skewness of the surface height and more frequent extreme events \cite{bolles2018anomalous}.
Here we develop a statistical dynamical model to explain and quantitatively
predict this anomalous behavior as experimental control
parameters are varied. The first step is to use incoming and outgoing
truncated Korteweg-de Vries (TKdV) equations matched in time at the
ADC. The TKdV equation is a Hamiltonian system which induces incoming
and outgoing statistical Gibbs invariant measures which are statistically
matched at the ADC. The statistical matching of the known nearly Gaussian
incoming Gibbs state at the ADC completely determines the predicted
anomalous outgoing Gibbs state, which can be calculated by a simple
MCMC algorithm, verified by direct numerical simulations, and successfully
captures key features of the experiment. There is even an analytic
formula for the anomalous outgoing skewness. The strategy here should
be useful for predicting extreme anomalous statistical behavior in
other dispersive media in different settings.

@@

\section{Experiments showing anomalous wave statistics induced by an abrupt depth change}

% ORIGINAL
%A series of experiments are carried out \cite{bolles2018anomalous}
%studying the anomalous statistical behavior of surface water waves
%crossing an abrupt depth transition. 
%The unidirectional waves propagate along a water tank over a step in the bottom topography,
%and the surface displacements of the wave levels are measured at several
%upstream and downstream locations. The wave field is excited by a
%paddle wheel forcing with angle

Controlled laboratory experiments were carried out in  \cite{bolles2018anomalous} to examine the statistical behavior of surface waves crossing an ADC. In these experiments, nearly unidirectional waves are generated by a paddle wheel and propagate through a long, narrow wave tank. Midway through, the waves encounter a step in the bottom topography, and thus abruptly transition from one depth to another. The paddle wheel is forced with angle
\[
\theta\left(t\right)=\theta_{0}+\Delta\theta\sum_{n=1}^{N}a_{n}\cos\left(\omega_{n}t+\delta_{n}\right),\;E\sim\left(\Delta\theta\right)^{2}\sum_{n}a_{n}^{2}\omega_{n}^{2}.
\]
where the weights $a_n$ are Gaussian in spectral space, and
the total energy $E$ in the system is determined by the angle amplitude
$\Delta\theta$. Optical measurements of the free surface reveal a number of surprising statistical features:
\begin{itemize}
\item Distinct statistics are found between the incoming and outgoing wave
disturbances: the incoming waves display near-Gaussian statistics,
while the outgoing waves show skewness toward the positive displacement.
\item The non-Gaussian statistics is related with the total energy contained
in the system: larger driving amplitudes $\Delta\theta$ generate
stronger skewness in the outgoing PDFs.
\item The waves also show different characteristic peak wavelengths in incoming
and outgoing flows.

\todo[inline]{I am not sure we need to include this last item. It is not really an experimental observation, simply a calculation that can be done using the dispersion relation and the two depths. Could we simply take this last item out?}
